---
title: 'Assignment 5: Data Visualization'
author: "Rosalind Hu"
date: "Fall 2024"
output:
  pdf_document: default
  html_document:
    df_print: paged
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options:
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory.
library(tidyverse)
library(lubridate)
library(here)
library(cowplot)
library(ggthemes)

rm(list=ls())
getwd()
here()
PeterPaul.chem.nutrients <- 
  read.csv(here("Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"), stringsAsFactors = TRUE)
NiwotRidge.litter <-
  read.csv(here("Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"), stringsAsFactors = TRUE)

#2 Make sure R is reading dates as date format; and change the format to date 
  #by using lubridate
# using glimpse(PeterPaul.chem.nutrients) to check the date format
PeterPaul.chem.nutrients$sampledate <- ymd(PeterPaul.chem.nutrients$sampledate)
NiwotRidge.litter # Question: how could I use function to examine each data property, like how to check Sampledate as Date format
NiwotRidge.litter$collectDate <- ymd(NiwotRidge.litter$collectDate)
```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3 Build a theme and set it as the default theme
Beige_theme <-
  theme(
    axis.title = element_text(color = "darkblue", face = "bold", family = "Times New Roman"),
    legend.position = "bottom",
    panel.background = element_rect(fill = "beige")
  )

theme_set(Beige_theme) #Once we use the theme_set, then the graphs we run 
                #afterwards will be applied as default
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 Plot total phosphorus('tp_ug') by phosphate('po4') with seperate aesthetics 
#  for Peter and Paul lakes
  
second_plot <- ggplot(PeterPaul.chem.nutrients, aes(x = po4, y = tp_ug, color = lakename)) +
  geom_point() +  
  #using lm method to generate linear model
  geom_smooth( method = "lm") + 
  xlim(0,50)+
  ylim(0,150)
  
print(second_plot)

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 Make a boxplot of temperature with month on x-axis

Temperature_Boxplot <- ggplot(PeterPaul.chem.nutrients,
                               aes(x = factor(month, levels = 1:12, labels = month.abb),
                                   y = temperature_C)) +
  geom_boxplot(aes(color = lakename)) +
  theme_minimal() +
  xlab("") +
  theme(legend.position = "top")

Temperature_Boxplot


# Make a boxplot of TP

TP_Boxplot <-
  ggplot(PeterPaul.chem.nutrients, 
         aes(x =factor(month, levels = 1:12,labels = month.abb), show.legend = F, y = tp_ug)) +
  geom_boxplot(aes(color = lakename))  +
theme_minimal() +
    xlab("") +
  theme(legend.position = "none")

TP_Boxplot

# Make a boxplot of TN

TN_Boxplot <-
  ggplot(PeterPaul.chem.nutrients,
         aes(x =factor(month, levels = 1:12,labels = month.abb), show.legend = F, y = tn_ug)) +
  geom_boxplot(aes(color = lakename)) +
theme_minimal() +
  xlab("Month")+
  theme(legend.position = "none")

TN_Boxplot

# Using cowplot to combine three plots
combined_plot <- plot_grid(
  Temperature_Boxplot, TP_Boxplot, TN_Boxplot,  
  ncol = 1,  # Stack plots vertically
  align = "v",  # Align the plots vertically for proper axis alignment
  rel_heights = c(1.4,1,1)
)
print(combined_plot)

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: In temrs of seasons, increasing from February to summer, peaking 
#between July and September, and then declining in the fall. Peter Lake 
consistently shows slightly higher median temperatures than Paul Lake during the
#summer. For total phosphorus and total nitrogen, concentrations are higher and 
#more variable during the warmer months (May to September), with Peter Lake 
#generally exhibiting greater variability and higher levels of both nutrients 
#compared to Paul Lake. These seasonal trends and inter-lake differences suggest
#a possible link between temperature and nutrient cycling, with Peter Lake 
#experiencing more nutrient loading, especially during summer.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
# Plot a subset of the litter dataset to include only the "Needles" functional group
Needles_data <- NiwotRidge.litter %>% 
  filter(functionalGroup == "Needles")

Needles_Plot <- ggplot(Needles_data, 
                       aes(x = collectDate, y = dryMass, color = nlcdClass)) +
  geom_point() +  # Using points to display dry mass by date
  geom_line() +   # Adding lines to connect points over time
  theme_minimal() +
  labs(title = "Dry Mass of Needle Litter by Date",
       x = "collectDate",
       y = "dryMass") +
  theme(legend.position = "top")

print(Needles_Plot)


#7 seperate three factors nlcdClass instead of using color function

Needles_Plot <- ggplot(Needles_data, 
                       aes(x = collectDate, y = dryMass, color = nlcdClass)) +
  geom_point() +  
  geom_line() +  
  theme_minimal() +
  labs(title = "Dry Mass of Needle Litter by Date",
       x = "Collect Date",
       y = "Dry Mass (g)") +
  scale_color_manual(values = c("evergreenForest" = "pink", 
                                "grasslandHerbaceous" = "green", 
                                "shrubScrub" = "blue")) +  
  theme(legend.position = "top")

print(Needles_Plot)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: Obviously the function color is more effective as I do not need to 
# create more lines to set each catergories into differnt color.